Category-based and location-based volitional covert attention affect memory at different timescales

### 2023-03-25 paper/main.tex
finished updating draft of text through end of methods section

Here we examine the ways different types of volitional covert attention
interact to affect memory. We designed an experimental
paradigm~\citep[following][]{Posn80} that asked participants to attend to a
series of presented composite image pairs while keeping their gaze fixed on a
central point. The image pairs comprised a left and right image, each
constructed by blending an image of a face and scene. The stimuli and
presentation durations were constant across the two experiments, but the
experiments differed in how often we asked participants to change the focus of
their attention with respect to image category (face versus scene) and image
location (left versus right). After the participants attended to a series of
images, we used a recognition memory test to assess which aspects of the
presented images had been encoded into memory. In both experiments we found
that the images participants covertly attended to were better recognized than
other images, supporting the notion that attention enhances memory
encoding~\citep[i.e., they rated attended images as more familiar than
unattended images][]{Yone02}. After maintaining the focus of attention to a
single image category and location (Sustained Attention Experiment),
participants also recognized the attended-category image at the unattended
location, and (to a lesser extent) the unattended-category image at the
attended location. After more rapidly varying their focus of attention
(Variable Attention Experiment), participants showed a similar boost in
recognition for the unattended-category image at the attended location, but
they did not recognize images at the unattended location. This suggests that
participants were able to shift the location of their covert attentional focus
more rapidly than they were able to shift their focus of covert attention to
stimulus features. We also found differences in the timecourses of these memory
effects, suggesting that the impact of location-based attention on memory
persists on the order of several seconds longer than the impact of
feature-based attention.

@@

@@

\begin{figure*}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figures/paradigm_and_key.pdf}

  \caption{\textbf{Paradigm overview.} \textbf{a.} Composite face/scene image.
  \textbf{b.} A single pair of composite images and a central fixation cross.
  \textbf{c.} Timecourse of an attention cue practice trial (3 total).
  \textbf{d.} Timecourse of a practice trial for the reaction time task (3
  total). \textbf{e.} Familiarity judgement practice trial (4 total).
  \textbf{f.} Timecourse of one Sustained Attention Experiment presentation
  phase and memory phase block. \textbf{g.} Timecourse of one Variable
  Attention Experiment presentation phase and memory phase block. \textbf{h.}
  Breakdown of presented (old) and novel stimuli that matched (or did not
  match) the focus of participants' feature-based attention and/or
  location-based attention. Filled colored circles on the right denote that the
  given stimulus (matching the corresponding color; images shown on the left)
  satisfied the indicated property. Open circles denote that the given stimulus
  did \textit{not} satisfy the indicated property. Gray circles denote
  properties that are undefined for the given stimulus.}

\subsection*{Participants}
\subsubsection*{Sustained Attention Experiment}
Sixty students, enrolled in one or more undergraduate psychology courses at Dartmouth College, participated in the Sustained Attention Experiment for course credit.  The participants' ages ranged from 18--21; 42 participants were female, 17 were male, and one identified as gender non-binary.  All participants had self-reported normal or corrected-to-normal vision, memory, and attention.  Participants gave written consent to enroll in the study under a protocol approved by the Committee for the Protection of Human Subjects at Dartmouth College.

\subsubsection*{Variable Attention Experiment}
Fifty-three individuals participated in the Variable Attention Experiment for either cash ($n=13$) or course credit ($n=40$; students enrolled in one or more undergraduate psychology courses at Dartmouth College).  The participants' ages ranged from 18--23; 34 participants were female and 19 were male.  All participants had self-reported normal or corrected-to-normal vision, memory, and attention.  Participants gave written consent to enroll in the study under a protocol approved by the Committee for the Protection of Human Subjects at Dartmouth College.

\subsection*{Stimulus selection and presentation}
Participants viewed photographs of faces, scenes, and composite images each comprising an equal blend of one face image and one scene image.  The pool of 360 face images included photographs of adult human male and female faces selected from the FERET database~\citep{PhilEtal98}.  The pool of 360 scene images included photographs of indoor and outdoor scenes selected from the SUN database~\citep{XiaoEtal10}.  The images we used from both databases came from a stimulus subset that was manually curated by Megan deBettencourt (personal communication).  All images were resized to $256 \times 256$ pixels, converted to greyscale, and processed so that every image was matched for mean contrast and intensity.  We selected 20 face images and 20 scene images from the stimulus pool to use in the instructional and practice phases of the experiments (Fig.~\ref{fig:paradigm}a--e).

In addition to the face and scene images, we presented (in white) attention cues to direct the participant's focus of attention.  The attention cues comprised a stylized icon of a face or mountain peaks, directing attention to the face or scene component of the images, respectively; and a left- or right-facing angled bracket, directing attention to the left or right image, respectively (e.g., Figs.~\ref{fig:paradigm}c, d, f, and g).

 Both experiments were conducted in a sound- and light-attenuated testing room containing a chair, desk, and 27-inch iMac desktop computer (display resolution: $2048 \times 1152$).  The participant sat in the chair and rested their chin on a chin rest located 60~cm from the display.  The active portion of the display screen occupied 52.96$^\circ$ (width) and 31.28$^\circ$ (height) of the participant's field of view from the chin rest.  Stimuli were sized to occupy $6.7^\circ$ (width and height) of the participant's field of view from the chin rest.  We maintained a black background (with any text displayed in white) throughout the experiment.

\subsection*{Eyetracking}
We recorded participants' eye gaze positions using a desk-mounted video-based eyetracker with a spatial resolution of 0.1$^\circ$ visual angle root mean squared error and a sampling rate of 30~Hz (Eye Tribe, The Eye Tribe, Copenhagen, Denmark).  We calibrated the eyetracker using a 9-point gaze pattern.  As described below, we re-calibrated the eyetracker at regular intervals throughout the experiments to protect against camera drift.

\subsection*{Experimental paradigm}
Both experimental paradigms comprised a practice phase followed by a series of eight task blocks.  Each task block was in turn comprised of a presentation phase and a memory phase.  The practice and presentation phases differed across the two experiments, and the memory phases were identical across the two experiments.  Both experiments were implemented using PsychoPy~\citep{PeirEtal19}.

\subsubsection*{Practice phase}
Several participants in pilot versions of our experiments reported that they found it difficult to modulate the focus of their attention quickly on command.  We therefore designed a practice sequence to orient the participant to the process of quickly modulating the focus of their attention without moving their eyes.  The experimenter remained in the testing room throughout the practice phase and answered any questions about the experiment.  The practice sequence builds up incrementally to provide a gradual on-ramping for the participant prior to beginning the main experimental tasks that we focused on in our analyses.

\paragraph{Practice shifting the focus of feature-based attention to elements of a single composite image.}
At the start of the practice phase, we instructed the participant to look at a single composite (face-scene blend) image at the center of the screen, and to try to bring the face component of the image into greater focus by attending to it (Fig.~\ref{fig:paradigm}a).  After pressing a button on the keyboard to indicate that they had done so, we displayed a second composite image and instructed the participant to bring the scene component of the new composite image into focus.  Again, they pressed a button to indicate that they had done so.

\paragraph{Practice shifting the focus of feature-based and location-based attention while viewing two composite images.}
Next, we asked the participant to stare at a fixation cross presented in the center of the screen while two composite images were displayed on the left and right side of the screen, respectively.  We first instructed the participant to attend to the scene component of the left image without moving their eyes.  Participants practiced shifting their attention, and they pressed a button on the keyboard to indicate that they had done so. We then displayed a second pair of composite images and instructed the participant to attend to the face component of the right image. Again, the participant shifted their attention in a self-paced manner, and pressed a button to indicate when they had successfully done so (Fig.~\ref{fig:paradigm}b).

\paragraph{Practice sustaining feature-based and location-based attention over a series of composite image pairs.}
We asked participants in the Sustained Attention Experiment to practice holding their focus of feature-based and location-based attention constant (to the face component of the right image) while viewing a series of three composite image pairs presented in succession (Fig.~\ref{fig:paradigm}c).

\paragraph{Practice varying feature-based and location-based attention over a series of composite image pairs.}
We asked participants in the Variable Attention Experiment to practice varying their focus of feature-based and location-based attention while viewing a series of three composite image pairs presented in succession (Fig.~\ref{fig:paradigm}c).

\paragraph{Practice reaction time probe.}
After practicing modulating their focus of attention to a series of composite image pairs, we introduced a reaction time probe after each image presentation, whereby we presented either an $\times$ or $\circ$ on either the left or the right of the screen (Fig.~\ref{fig:paradigm}d).  We asked the participant to press the \texttt{1} key as quickly as possible when they saw an $\times$, or the \texttt{3} key as quickly as possible when they saw an $\circ$.  We did not impose a time limit on their responses, other than asking participants to respond as quickly as they were able.  Participants practiced three trials of modulating their focus of attention to a pair of composite images (3~s), and reacting as quickly as possible to the $\times$ or $\circ$ symbol presented after each composite image pair.  The reaction time probe was intended to keep participants continually engaged in modulating the focus of their attention.

\paragraph{Practice recognition memory task.}
Finally, we asked the participant to practice reporting familiarity on a recognition memory task (Fig.~\ref{fig:paradigm}e).  We presented a single face or scene image at the center of the screen, and asked them to press a button to indicate how ``familiar'' the image seemed: \texttt{1} (very confident they had not seen the image), \texttt{2} (somewhat confident they had not seen the image), \texttt{3} (somewhat confident they had seen the image), or \texttt{4} (very confident that they had seen the image).  We instructed the participant to go with their ``gut reaction'' in the event that they were unsure of how to respond.  We allowed the participant up to 2~s to provide their response.  We gave participants a total of four practice images to rate.

\subsubsection*{Task blocks}
During each task block we asked the participant to modulate their attention while viewing a series of 10 composite image pairs (each followed by a reaction time probe), and then we tested the participant's memory using 40 familiarity judgements.  Each participant completed a total of eight task blocks.

\paragraph*{Sustained Attention Experiment: presentation phase.} Participants viewed an attention cue (1.5~s) instructing them to attend to either the face or scene component of either the left or right images in each to-be-viewed composite pair.  Next we displayed 10 composite images in succession (each preceded by a fixation cross and proceeded by a reaction time probe).  Complete timing information is displayed in Figure~\ref{fig:paradigm}f.  All possible attention cue pairs appeared exactly twice across the eight task blocks.

\paragraph*{Variable Attention Experiment: presentation phase.}  Participants viewed a succession of 10 attention cues (1.5~s), each followed by a fixation cross (1~s), composite image (3~s), and a reaction time probe (Fig.~\ref{fig:paradigm}g).  The attention cues were selected pseudorandomly across trials within each block, with the constraints that no single attention cue pair could appear more than three times across the 10 composite image pairs within a single task block.

\paragraph*{Memory phase.} After the presentation phase of each task block, we asked the participant to rate the familiarity (on a 1--4 scale, as during the practice phase) of a succession of face and scene images.  Each image was preceded by a 1~s fixation cross, and participants had up to 2~s to input their rating of each image.  Participants made a total of 40 familiarity judgements, about 20 face images and 20 scene images.  Of these, 20 of the images (10 faces and 10 scenes) were drawn randomly from the (attended and unattended) composite images that the participant had viewed during the presentation phase.  The remaining 20 images (10 faces and 10 scenes) were novel images that the participant had not encountered during any part of the experiment.  At the end of each memory block, the participant was given the opportunity to take a short break.  When they were ready to continue with the next task block, they indicated their readiness to the experimenter.  The experimenter then entered the testing room, re-ran the eyetracker calibration sequence, and exited the testing room prior to the next task block.

\section*{Results}
We ran two volitional covert attention experiments; in the Sustained Attention Experiment we asked participants to \textit{sustain} the focus of their attention over a succession of 10 stimulus presentations per block whereas in the Variable Attention Experiment we asked participants to \textit{vary} their focus of attention with each new stimulus (also for a total of 10 stimulus presentations per block).  Each stimulus comprised a pair of composite images (one on the left and one on the right side of the display), where each composite comprised an equal blend of a unique face and a unique scene image.  We followed the presentation phases of each experimental block with a memory phase, where participants performed a recognition memory task by rating the familiarity of previously experienced and novel face and scene images (see \textit{Materials and methods}, Fig.~\ref{fig:paradigm}).

We considered each image we probed in the memory phase with respect to whether (and how) the participant had experienced and/or attended to that image during the presentation phase (Fig.~\ref{fig:familiarity}).  For example, if a given face image had been presented as part of the composite image on the right of the screen following an \texttt{attend to the face on the right} instruction, then that face image would have been the focus of the participant's attention during one stimulus presentation.  The face image on the left (unattended side) during that trial would have been captured by feature-based attention, but not by location-based attention; and the scene image (unattended category) on the right during that trial would have been captured by location-based attention, but not by feature-based attention.  The scene image on the left during that trial remained outside of the focus of both feature-based and location-based attention.  In this way, we categorized each of the images that participants experienced during the presentation phases of each experiment in terms of whether they fell under the scope of feature-based and/or location-based attention.  We also categorized novel face and scene images that we asked participants to judge during the memory phase (i.e., images that the participant hadn't seen before) as belonging to the attention-cued or uncued category.  These novel images were intended to serve as a baseline for comparison.  For example, we sought to measure how people rated the familiarity of new images in general.  We also used differences in familiarity ratings between novel images of the attention-cued versus uncued categories to evaluate potential \textit{prospective} affects of modulating the focus of feature-based attention.

\begin{figure*}[tp]
  \centering
  \includegraphics[width=0.75\textwidth]{figures/familiarity}
  \caption{\textbf{Familiarity ratings for attended, unattended, and novel stimuli.}  Each split violin plot displays the distribution of within-participant average familiarity ratings given to faces (left, darker colors) and scenes (right, lighter colors) during the memory phases of each experiment.  As shown in the legend (bottom), the colors indicate whether each image had been viewed during the presentation phase (\texttt{old}) or not; whether the given images matched the \texttt{attended category}; and/or whether the given images matched the \texttt{attended location}.  The colored lines above each set of violin plots denote statistical differences  (\textbf{pos}itive or \textbf{neg}ative differences in mean, collapsing over image category, assessed via two-tailed $t$-tests) between the distributions centered on the endpoints of each line.  The line thicknesses denote $p$-values as indicated in the legend.  Asterisks denote differences between the face versus scene distributions (assessed via two-tailed $t$-tests).  Panel \textbf{a.} displays results from the Sustained Attention Experiment and panel \textbf{b.} displays results from the Variable Attention Experiment.  Figure~\familiarity~displays these results broken down by participant cohort.}
\label{fig:familiarity}
\end{figure*}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figures/exps1_2_xgaze.pdf}
  \caption{\textbf{Horizontal coordinates of visual fixations.}  Each violin plot displays a distribution of participant-wise average horizontal gaze positions relative to a presented image.  The coordinates have been normalized such that a value of 1.0 denotes the furthest onscreen coordinate from the central fixation point \textit{towards} the direction of the given image, and -1.0 denotes the furthest onscreen coordinate from the central fixation point \textit{away from} the direction of the given image.  The gray bars in each panel mark the boundaries of the presented images.  The violin plots are broken down by participants' familiarity ratings during the memory phases of each experiment, and by each image's relation to the attention cue while that image appeared onscreen.  Panel \textbf{a.} displays fixation data for the Sustained Attention Experiment and Panel \textbf{b.} displays fixation data for the Variable Attention Experiment.  Figure~\fixation~displays these results broken down by participant cohort.}
\label{fig:fixation}
\end{figure}

@@

\begin{figure*}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figures/Cat_Cued_TimeSeries.pdf}
  \caption{\textbf{Familiarity ratings over time for images that matched the cued or uncued image category.}  Each curve reflects the average familiarity ratings for attended, unattended, and novel images (denoted in the legends on the right) within a succession of overlapping 20-image sliding windows.  Error ribbons denote 95\% confidence intervals, computed across participants.  Panels \textbf{a.} and \textbf{c.} display results from the Sustained Attention Experiment and Panels \textbf{b.} and \textbf{d.} display results from the Variable Attention Experiment.  The paired horizontal lines at the bottom of each panel denote timepoints when the given pair of curves was statistically distinguishable (i.e., the topmost line color was statistically greater than the bottommost line color at $\alpha = 0.05$, via a paired two-tailed $t$-test.)  The gray lines in Panels b and d reflect familiarity ratings of novel stimuli from both the most recently attended and unattended image categories.   Figures~\timecourseCuedCategory~and \timecourseUncuedCategory~display these results broken down by participant cohort.}
\label{fig:familiarity_timecourse_category}
\end{figure*}

@@

\begin{figure*}[tp]
  \centering
  \includegraphics[width=\textwidth]{figures/Side_Cued_TimeSeries_2.pdf}
  \caption{\textbf{Familiarity ratings over time for images that matched the cued or uncued image location.}  This figure follows the same format and conventions as Figure~\ref{fig:familiarity_timecourse_category}.   Figures~\timecourseCuedLocation~and \timecourseUncuedLocation~display these results broken down by participant cohort.}
\label{fig:familiarity_timecourse_location}
\end{figure*}

@@

J. R. Manning and K. Ziman developed the concept for this study. Experiment code was written by K. Ziman and A. R. Martinez, and testing and data collection were conducted by M. R. Lee and K. Ziman. K. Ziman, M. R. Lee, and A. R. Martinez analyzed the data.  All authors contributed to writing the manuscript and have approved the final version of the manuscript for submission.

>>>>

Here we examine the ways different types of volitional covert attention
interact to affect memory. We designed an experimental
paradigm~\citep[following][]{Posn80} that asked participants to attend to a
series of presented composite image pairs while keeping their gaze fixed on a
central point. The image pairs comprised a left and right image, each
constructed by blending an image of a face and place. The stimuli and
presentation durations were constant across the two experiments, but the
experiments differed in how often we asked participants to change the focus of
their attention with respect to image category (face versus place) and image
location (left versus right). After the participants attended to a series of
images, we used a recognition memory test to assess which aspects of the
presented images had been encoded into memory. In both experiments we found
that the images participants covertly attended to were better recognized than
other images, supporting the notion that attention enhances memory
encoding~\citep[i.e., they rated attended images as more familiar than
unattended images][]{Yone02}. After maintaining the focus of attention to a
single image category and location (Sustained Attention Experiment),
participants also recognized the attended-category image at the unattended
location, and (to a lesser extent) the unattended-category image at the
attended location. After more rapidly varying their focus of attention
(Variable Attention Experiment), participants showed a similar boost in
recognition for the unattended-category image at the attended location, but
they did not recognize images at the unattended location. This suggests that
participants were able to shift the location of their covert attentional focus
more rapidly than they were able to shift their focus of covert attention to
stimulus features. We also found differences in the timecourses of these memory
effects, suggesting that the impact of location-based attention on memory
persists on the order of several seconds longer than the impact of
feature-based attention.

@@

\begin{figure*}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/paradigm}

  \caption{\textbf{Experimental paradigm.} \textbf{A.--D. Practice phase.}
  \textbf{A.} Composite face/place image. \textbf{B.} A single pair of
  composite images and a central fixation cross. \textbf{C.} One attention cue
  practice trial. \textbf{D.} Familiarity judgement practice trial. \textbf{E.
  Sustained attention condition.} Participants receive an attention cue,
  followed by a sequence of 10 composite image pairs (Panel E), and then they
  make of familiarity judgements on each of 40 face and place images, presented
  in sequence (Panel F).  \textbf{G.  Variable attention condition.}  Participants
  study a succession of 10 composite image pairs, each preceded by an attention cue.
  After studying the images, they make a series of 40 familiarity judgements about
  face and place images, as in the sustained attention condition (Panel F).}

\subsection*{Participants}

A total of 53 Dartmouth College undergraduate students enrolled in our study
(30 in the sustained attention condition and 23 in the variable attention
condition). Following a pilot study using a similar experimental design, we
aimed to enroll 30 participants in each condition. However, we fell short of
our enrollment target in the variable attention condition when in-person
testing was discontinued at our institution due to the COVID-19 pandemic. All
participants had self-reported normal or corrected-to-normal vision, memory,
and attention. Participants gave written consent to enroll in the study under a
protocol approved by the Committee for the Protection of Human Subjects at
Dartmouth College.

We used a voluntary pre-experimental survey to collect self-reported
demographic information about each partipant. All 53 participants elected to
fill out the survey. Participants ranged in age from 18--21 years old (mean:
18.7 years; standard deviation: 0.8 years). Participants reported their genders
as female (34 particiapnts), male (18 participants), and gender non-binary (1
participant). Participants reported their ethnicities as Not Hispanic or Latino
(44 particiapnts), Hispanic or Latino (7 participants), or declined to respond
(2 participant). Participants reported their race as White (37 participants),
Asian (13 participants), American Indian or Alaska Native (4 participants),
Black or African American (2 participants), and Other (1 participant). Note
that each participant could report one or more racial categories, as they
deemed appropriate.

Fourty-nine participants reported having no reading impairments, and 4
participants reported having reading impairments such as dyslexia. Fifty
participants reported having normal color vision and 3 reported having abnormal
color vision such as colorblindness. Fifty participants reported taking no
medications and having no recent injuries. One participant reported that they
had recently ``hit [their] head very hard.'' Another reported having taken
concerta in the past, but mentioned they had not taken it recently. One
participant reported using amphetamines regularly, but also clarified that they
were not currently on amphetamines at the time of testing.

We also asked participants to self-report on their sleep, alertness, and coffee
consumption. Participants reported having gotten between 4 and 9 hours of sleep
on the night prior to testing (mean: 6.9 hours; standard deviation: 1.3 hours).
Participants reported their alertness at the time of testing, and we converted
their responses to point values as follows: ``Very alert'' (5 points), ``A
little alert'' (4 points), ``Neutral'' (3 points), ``A little sluggish'' (2
points), and ``Very sluggish'' (1 point). Across all participants, the full
range of alertness values were used (maximum: 5; minimum: 1; mean: 3.44;
standard deviation: 1.0). Participants reported having consumed between 0 and 2
cups of coffee so far on their testing day (mean: 0.3 cups; standard deviation:
0.5 cups).

\subsection*{Stimulus selection and presentation}

Participants viewed photographs of faces, places, and composite images each
comprising an equal blend of one face image and one place image. The pool of
360 face images included photographs of adult human male and female faces
selected from the FERET database~\citep{PhilEtal98}. The pool of 360 place
images included photographs of indoor and outdoor places selected from the SUN
database~\citep{XiaoEtal10}. The images we used from both databases came from a
stimulus subset that was manually curated by Megan deBettencourt (personal
communication). All images were resized to $256 \times 256$ pixels, converted
to greyscale, and processed so that every image was matched for mean contrast
and intensity. We selected 20 face images and 20 place images from the stimulus
pool to use in the instructional and practice phases of the experiments
(Fig.~\ref{fig:paradigm}A--D).

In addition to the face and place images, we presented (in white) attention
cues to direct the participant's focus of attention. The attention cues
comprised a stylized icon of a face or mountain peaks, directing attention to
the face or place component of the images, respectively; and a left- or
right-facing angled bracket, directing attention to the left or right image,
respectively (e.g., Figs.~\ref{fig:paradigm}C, E, and G).

 Our experiment was conducted in a sound- and light-attenuated testing room
 containing a chair, desk, and 27-inch iMac desktop computer (display
 resolution: $2048 \times 1152$). The participant sat in the chair and rested
 their chin on a chin rest located 60~cm from the display. The active portion
 of the display screen occupied 52.96$^\circ$ (width) and 31.28$^\circ$
 (height) of the participant's field of view from the chin rest. Stimuli were
 sized to occupy $6.7^\circ$ (width and height) of the participant's field of
 view from the chin rest. We maintained a black background (with any text
 displayed in white) throughout the experiment.

\subsection*{Eyetracking}

We recorded participants' eye gaze positions using a desk-mounted video-based
eyetracker with a spatial resolution of 0.1$^\circ$ visual angle root mean
squared error and a sampling rate of 30~Hz (Eye Tribe, The Eye Tribe,
Copenhagen, Denmark). We calibrated the eyetracker using a 9-point gaze
pattern. As described below, we re-calibrated the eyetracker at regular
intervals throughout the experiment to protect against camera drift.

\subsection*{Experimental paradigm}

Our experiment comprised two testing conditions: a \textit{sustained} attention
condition and a \textit{variable} attention condition. Both experimental
conditions comprised a practice phase followed by a series of eight task
blocks. Each task block was in turn comprised of a presentation phase and a
memory phase. The practice and presentation phases differed across the two
experiments, and the memory phases were identical across the two experiments.
Both experiments were implemented using PsychoPy~\citep{PeirEtal19}.

\subsubsection*{Practice phase}

Several participants in pilot versions of our experiments reported that they
found it difficult to modulate the focus of their attention quickly on command.
We therefore designed a practice sequence to orient the participant to the
process of quickly modulating the focus of their attention without moving their
eyes. The experimenter remained in the testing room throughout the practice
phase and answered any questions about the experiment. The practice sequence
builds up incrementally to provide a gradual on-ramping for the participant
prior to beginning the main experimental tasks that we focused on in our
analyses.

\paragraph{Practice shifting the focus of category-based attention to elements
of a single composite image.}

At the start of the practice phase, we instructed the participant to look at a
single composite (face-place blend) image at the center of the screen, and to
try to bring the face component of the image into greater focus by attending to
it (Fig.~\ref{fig:paradigm}A). After pressing a button on the keyboard to
indicate that they had done so, we displayed a second composite image and
instructed the participant to bring the place component of the new composite
image into focus. Again, they pressed a button to indicate that they had done
so.

\paragraph{Practice shifting the focus of category-based and location-based
attention while viewing two composite images.}

Next, we asked the participant to stare at a fixation cross presented in the
center of the screen while two composite images were displayed on the left and
right side of the screen, respectively (Fig.~\ref{fig:paradigm}B). We first
instructed the participant to attend to the place component of the left image
without moving their eyes. Participants practiced shifting their attention, and
they pressed a button on the keyboard to indicate that they had done so. We
then displayed a second pair of composite images and instructed the participant
to attend to the face component of the right image. Again, the participant
shifted their attention in a self-paced manner, and pressed a button to
indicate when they had successfully done so.

\paragraph{Practice \textit{sustaining} category-based and location-based attention over
a series of composite image pairs.}

We asked participants in the sustained attention condition to practice holding
their focus of category-based and location-based attention constant (to the
face component of the right image) while viewing a series of three composite
image pairs presented in succession (Fig.~\ref{fig:paradigm}C).

\paragraph{Practice \textit{varying} category-based and location-based
attention over a series of composite image pairs.}

We asked participants in the variable attention condition to practice varying
their focus of category-based and location-based attention while viewing a
series of three composite image pairs, each presented after a different
attention cue (Fig.~\ref{fig:paradigm}C).

\paragraph{Practice reaction time probe.}

After practicing modulating their focus of attention to a series of composite
image pairs, we introduced a reaction time probe after each image presentation,
whereby we presented either an $\times$ or $\circ$ on either the left or the
right of the screen (not shown). We asked the participant to press the
\texttt{1} key as quickly as possible when they saw an $\times$, or the
\texttt{3} key as quickly as possible when they saw an $\circ$. We did not
impose a time limit on their responses, other than asking participants to
respond as quickly as they were able. Participants practiced three trials of
modulating their focus of attention to a pair of composite images (3~s), and
reacting as quickly as possible to the $\times$ or $\circ$ symbol presented
after each composite image pair. The reaction time probe was intended to keep
participants continually engaged in modulating the focus of their attention.

\paragraph{Practice recognition memory task.}

Finally, we asked the participant to practice reporting familiarity on a
recognition memory task (Fig.~\ref{fig:paradigm}D). We presented a single face
or place image at the center of the screen, and asked them to press a button to
indicate how ``familiar'' the image seemed: \texttt{1} (very confident they had
not seen the image), \texttt{2} (somewhat confident they had not seen the
image), \texttt{3} (somewhat confident they had seen the image), or \texttt{4}
(very confident that they had seen the image). We instructed the participant to
go with their ``gut reaction'' in the event that they were unsure of how to
respond. We allowed the participant up to 2~s to provide their response. We
gave participants a total of four practice images to rate.

\subsubsection*{Task blocks}

During each task block we asked the participant to modulate their attention
while viewing a series of 10 composite image pairs (each followed by a reaction
time probe), and then we tested the participant's memory using 40 familiarity
judgements. Each participant completed a total of eight task blocks.

\paragraph*{Sustained attention condition: presentation phase
(Fig.~\ref{fig:paradigm}E).}

Participants viewed an attention cue (1.5~s) instructing them to attend to
either the face or place component of either the left or right images in each
to-be-viewed composite pair. Next we displayed 10 composite images in
succession (each preceded by a fixation cross and proceeded by a reaction time
probe). All possible attention cue pairs appeared exactly twice across the
eight task blocks.

\paragraph*{Variable attention condition: presentation phase
(Fig.~\ref{fig:paradigm}G).}

Participants viewed a succession of 10 attention cues (1.5~s), each followed by
a fixation cross (1~s), composite image (3~s), and a reaction time probe. The
attention cues were selected pseudorandomly across trials within each block,
with the constraints that no single attention cue pair could appear more than
three times across the 10 composite image pairs within a single task block.

\paragraph*{Memory phase (Fig.~\ref{fig:paradigm}F).}

After the presentation phase of each task block, we asked the participant to
rate the familiarity (on a 1--4 scale, as during the practice phase) of a
succession of face and place images. Each image was preceded by a 1~s fixation
cross, and participants had up to 2~s to input their rating of each image.
Participants made a total of 40 familiarity judgements, about 20 face images
and 20 place images. Of these, 20 of the images (10 faces and 10 places) were
drawn randomly from the (attended and unattended) composite images that the
participant had viewed during the presentation phase. The remaining 20 images
(10 faces and 10 places) were novel images that the participant had not
encountered during any part of the experiment. At the end of each memory block,
the participant was given the opportunity to take a short break. When they were
ready to continue with the next task block, they indicated their readiness to
the experimenter. The experimenter then entered the testing room, re-ran the
eyetracker calibration sequence, and exited the testing room prior to the next
task block.

@@

\section*{Results}

We ran two volitional covert attention experiments; in the Sustained Attention
Experiment we asked participants to \textit{sustain} the focus of their
attention over a succession of 10 stimulus presentations per block whereas in
the Variable Attention Experiment we asked participants to \textit{vary} their
focus of attention with each new stimulus (also for a total of 10 stimulus
presentations per block). Each stimulus comprised a pair of composite images
(one on the left and one on the right side of the display), where each
composite comprised an equal blend of a unique face and a unique place image.
We followed the presentation phases of each experimental block with a memory
phase, where participants performed a recognition memory task by rating the
familiarity of previously experienced and novel face and place images (see
\textit{Materials and methods}, Fig.~\ref{fig:paradigm}).

We considered each image we probed in the memory phase with respect to whether
(and how) the participant had experienced and/or attended to that image during
the presentation phase (Fig.~\ref{fig:familiarity}). For example, if a given
face image had been presented as part of the composite image on the right of
the screen following an \texttt{attend to the face on the right} instruction,
then that face image would have been the focus of the participant's attention
during one stimulus presentation. The face image on the left (unattended side)
during that trial would have been captured by feature-based attention, but not
by location-based attention; and the place image (unattended category) on the
right during that trial would have been captured by location-based attention,
but not by feature-based attention. The place image on the left during that
trial remained outside of the focus of both feature-based and location-based
attention. In this way, we categorized each of the images that participants
experienced during the presentation phases of each experiment in terms of
whether they fell under the scope of feature-based and/or location-based
attention. We also categorized novel face and place images that we asked
participants to judge during the memory phase (i.e., images that the
participant hadn't seen before) as belonging to the attention-cued or uncued
category. These novel images were intended to serve as a baseline for
comparison. For example, we sought to measure how people rated the familiarity
of new images in general. We also used differences in familiarity ratings
between novel images of the attention-cued versus uncued categories to evaluate
potential \textit{prospective} affects of modulating the focus of feature-based
attention.

\begin{figure*}[tp]
  \centering
  \includegraphics[width=0.75\textwidth]{figures/familiarity}

  \caption{\textbf{Familiarity ratings for attended, unattended, and novel
  stimuli.} Each split violin plot displays the distribution of
  within-participant average familiarity ratings given to faces (left, darker
  colors) and scenes (right, lighter colors) during the memory phases of each
  experiment. As shown in the legend (bottom), the colors indicate whether each
  image had been viewed during the presentation phase (\texttt{old}) or not;
  whether the given images matched the \texttt{attended category}; and/or
  whether the given images matched the \texttt{attended location}. The colored
  lines above each set of violin plots denote statistical differences
  (\textbf{pos}itive or \textbf{neg}ative differences in mean, collapsing over
  image category, assessed via two-tailed $t$-tests) between the distributions
  centered on the endpoints of each line. The line thicknesses denote
  $p$-values as indicated in the legend. Asterisks denote differences between
  the face versus scene distributions (assessed via two-tailed $t$-tests).
  Panel \textbf{a.} displays results from the Sustained Attention Experiment
  and panel \textbf{b.} displays results from the Variable Attention
  Experiment. Figure~\familiarity~displays these results broken down by
  participant cohort.}

\label{fig:familiarity}
\end{figure*}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{figures/exps1_2_xgaze.pdf}

  \caption{\textbf{Horizontal coordinates of visual fixations.} Each violin
  plot displays a distribution of participant-wise average horizontal gaze
  positions relative to a presented image. The coordinates have been normalized
  such that a value of 1.0 denotes the furthest onscreen coordinate from the
  central fixation point \textit{towards} the direction of the given image, and
  -1.0 denotes the furthest onscreen coordinate from the central fixation point
  \textit{away from} the direction of the given image. The gray bars in each
  panel mark the boundaries of the presented images. The violin plots are
  broken down by participants' familiarity ratings during the memory phases of
  each experiment, and by each image's relation to the attention cue while that
  image appeared onscreen. Panel \textbf{a.} displays fixation data for the
  Sustained Attention Experiment and Panel \textbf{b.} displays fixation data
  for the Variable Attention Experiment. Figure~\fixation~displays these
  results broken down by participant cohort.}

\label{fig:fixation}
\end{figure}

@@

\begin{figure*}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figures/Cat_Cued_TimeSeries.pdf}

  \caption{\textbf{Familiarity ratings over time for images that matched the
  cued or uncued image category.} Each curve reflects the average familiarity
  ratings for attended, unattended, and novel images (denoted in the legends on
  the right) within a succession of overlapping 20-image sliding windows. Error
  ribbons denote 95\% confidence intervals, computed across participants.
  Panels \textbf{a.} and \textbf{c.} display results from the Sustained
  Attention Experiment and Panels \textbf{b.} and \textbf{d.} display results
  from the Variable Attention Experiment. The paired horizontal lines at the
  bottom of each panel denote timepoints when the given pair of curves was
  statistically distinguishable (i.e., the topmost line color was statistically
  greater than the bottommost line color at $\alpha = 0.05$, via a paired
  two-tailed $t$-test.) The gray lines in Panels b and d reflect familiarity
  ratings of novel stimuli from both the most recently attended and unattended
  image categories. Figures~\timecourseCuedCategory~and
  \timecourseUncuedCategory~display these results broken down by participant
  cohort.}

\label{fig:familiarity_timecourse_category}
\end{figure*}

@@

\begin{figure*}[tp]
  \centering
  \includegraphics[width=\textwidth]{figures/Side_Cued_TimeSeries_2.pdf}

  \caption{\textbf{Familiarity ratings over time for images that matched the
  cued or uncued image location.} This figure follows the same format and
  conventions as Figure~\ref{fig:familiarity_timecourse_category}.
  Figures~\timecourseCuedLocation~and \timecourseUncuedLocation~display these
  results broken down by participant cohort.}

\label{fig:familiarity_timecourse_location}
\end{figure*}

@@

JRM and KZ developed the concept for this study. Experiment code was written by
KZ and ARM, and testing and data collection were conducted by MRL and KZ. KZ,
MRL, ARM, and JRM analyzed the data. JRM supervised the project. All authors
contributed to writing and editing the manuscript.

\section*{Data and code availability}
